\large{\bf Large effects and the infinitesimal model }

### 2023-07-05 writeup.tex
peter edits

% Intro:
%  - the infinitesimal model is useful because it is a "trait-only" model: it allows for predictions and modeling without knowing the underlying genetics.
%  - Its key properties are: (a) independence of seg noise, and (b) Gaussian-ness.
%  - It is basically a CLT, and in particular, (b) derives from a the CLT deriving from similar effect sizes.
%  - Motivating Q: there are other CLTs; can we generalize (b) to non-Gaussian effects, while keeping independence of seg variation?
% 
% Data motivation:
%  First, to see if this is worth asking we look at some data
%  ... and find that yeah, maybe effect sizes don't fall into the Gaussian domain of attraction.
% 
% First answer to the motivating question:  "No"
% Non-Gaussian effects (sometimes) imply nonindependence of segregation noise:
%   In at least some cases, natural non-gaussian models produce nonindependence:
%   1. all loci are uniformly small except there's one big one
%   2. loci are Cauchy
%   3. (is there a simple statement we can make more generally?)
% 
% Second answer, weaker but also "no":
%  - Observation: Let A and B be the contributions to parent phenotype, with A inherited by the offspring and B not. If A and B are independent, and seg noise is independent of parent phenotype, then A and B are Gaussian.
%  - When would A and B be asymptotically independent? One case is the rare-mutation limit that obtains a "thinned Poisson process" limit for A and B (I think: if all allele frequencies are at frequency O(1/M)?)
%  - (I have other thoughts to be sorted out here)
%  - Open Q (maybe?): is there any non-Gaussian model where segregation noise is independent of parental traits?
% 
% Simulations: let's see how this stuff works in practice
%  - neutral simulations
%  - also, stabilizing selection with theoretical motivation for form of stabilizing selection
%  - plot: (A) trait mean over time and trait distribution at the end
%  - plot: (B) (distrn of seg noise), (seg noise vs midparent), and (Radon-Nicodym deriv of conditional seg noise for a few values of midparent)
% 
% If we remove 'big loci' is what's left the infinitesimal model?
%  - people do this in practice (animal breeding papers)
%  - theory:  form of limiting conditional distribution
%  - theory: something about independence  (?? if we get there ??)
%  - plots (A) & (B) above from simulation, but for only small effects

@@

\begin{abstract}
    The infinitesimal model of quantitative genetics
    relies on the Central Limit Theorem to stipulate that
    under additive models of quantitative traits
    determined by many loci having similar effect size,
    the difference between an offspring's genetic trait component
    and the average of their two parents' genetic trait components
    is Normally distributed and independent of the parents' values.
    Here, we investigate how the assumption of similar effect sizes
    affects the model:
    in particular, if the tail of the effect size distribution is polynomial
    with exponent $\alpha < 2$, then sums of effects
    should be well-approximated by a ``stable distribution'',
    and find tail exponents between 1 and 2 in 
    effect sizes estimated by genome-wide association studies
    of many human disease-related traits.
    We show that the independence of offspring trait deviations from parental averages
    in many cases implies a Gaussian distribution,
    suggesting that non-Gaussian models of trait evolution must expicitly
    track the underlying genetics, at least for loci of large effect.
    We also characterize possible limiting trait distributions
    of the infinitesimal model with infinitely divisible noise distributions,
    and compare our results to simulations.
\end{abstract}

@@

Suppose that the focal individual's trait is
$$
    Z_M = \sum_{\ell=1}^M \eta_\ell / M^{1/\alpha},
$$
where $\eta_\ell$ is the contribution of the allele carried at the $\ell^\text{th}$ locus.
The effects of the alleles carried by the two parents
are $\eta_\ell^{[1]}$ and $\eta_\ell^{[2]}$, respectively,
and their traits are
$Z_M^{[1]} = \sum_\ell \eta_\ell^{[1]} / M^{1/\alpha}$ and
$Z_M^{[2]} = \sum_\ell \eta_\ell^{[2]} / M^{1/\alpha}$.
We assume the loci are unlinked and inheritance is fair, so
letting $X_\ell$ be a family of Bernoulli(1/2) random variables that are independent of everything else,
$$
    \eta_\ell = X_\ell \eta_\ell^{[1]} + (1 - X_\ell) \eta_\ell^{[2]} .
$$
The \emph{transmission noise}, $R_M$, is the offspring's trait minus the midparent value, $\bar Z_M$:
\begin{align*}
    \bar Z_M &= \frac{Z_M^{[1]} + Z_M^{[2]}}{2} \\
    R_M &= Z_M - \bar Z_M \\
        &= \sum_{\ell=1}^M
            \frac{
                X_\ell (\eta_\ell^{[1]} - \eta_\ell^{[2]})
                + (1 - X_\ell) (\eta_\ell^{[2]} - \eta_\ell^{[1]})
            }{ 2 } .
\end{align*}

@@

It is mathematically intriguing to speculate what effect a non-Gaussian distribution
for $\epsilon$ would have on the dynamics induced by~\eqref{eqn:offspring_model}.
But, is there any evidence that non-Gaussian segregation noise is important in practice?
In this paper we focus primarily on the theory,
but first take a brief and imperfect look at what some data have to say on the topic.
To do this, we downloaded the GWAS results provided by~\citet{biobankSNPs}
for 7,221 phenotypes from 361,194 individuals from the March 2018 release of the UK Biobank.
After filtering (see details below) we were left with 695 binary illness-related phenotypes.
Taken at face value, these provide for each phenotype a set of estimates of the additive effect
of the alternative allele at a set of SNPs on the phenotype relative to the reference allele.
The phenotypes we consider were all binary and coded as 0/1,
so estimates could be thought of as additive adjustments to the risk
(i.e., effects are \emph{not} on a logit scale).
These estimates were obtained as the coefficient for the SNP in a simple linear model
fit to each phenotype for each SNP,
with the first 20 principal components and all combinations of age, age squared, and inferred sex as covariates;
we considered effects of those SNPs with a reported $p$-value less than $10^{-8}$.
There are a great many possible issues with these data;
however, the results seem consistent across many SNPs and phenotypes,
and we know of no reason that possible methodological artifacts
(e.g., confounding by uncorrected variables)
would specifically affect the decay of the tail of effect sizes.

Suppose that the phenotype of each individual is additive:
the maternal trait value is $x = \sum_i x_{i,1} + x_{i,2}$
and the paternal trait value is $y = \sum_i y_{i,1} + y_{i,2}$,
where $x_{i,1}$ and $x_{i,2}$ are the effects of the alleles carried
by the mother on each of the two chromosomes
at the $i^\text{th}$ locus, respectively.
Then the offspring's trait is $z = \sum_i z_{i,m} + z_{i,p}$,
where $z_{i,m} = \theta_{i,m} x_{i,1} + (1-\theta_{i,m}) x_{i,2}$,
and $\theta_{i,m} = 1$ with probability 1/2 and $\theta_{i,m} = 0$ otherwise;
and likewise $z_{i,p} = \theta_{i,p} y_{i,1} + (1-\theta_{i,p}) y_{i,2}$.
With this notation, the segregation noise $z - \frac{x+y}{2}$
is a sum of terms of the form
$$
    \left( \theta_{i,m} - 1/2 \right) x_{i,1}
    + \left( 1/2 - \theta_{i,m} \right) x_{i,2} 
$$
(and similar terms with $y$ in place of $x$).
Now let $X$ be the effect of a uniformly chosen allele carried by a uniformly chosen individual
at a uniformly chosen locus.
The appropriate central limit theorem is determined by the tail behavior of $|(\theta - 1/2) X| = |X|/2$.
So, we'd like to know the value of
$$
    \lim_{t \to \infty} \frac{ \log \P\{ |X| > 2 t \} }{ \log t } = \alpha ,
$$
if the limit exists.
Let $p_i$ be the minor allele frequency at the $i^\text{th}$ locus;
since the data we have assigns effect 0 to the major allele,
$x_i = 0$ with probability $1-p_i$.
Therefore, if we have estimated effects for $N$ SNPs
with minor allele frequency $p_i$ and effect size $e_i$ for the minor allele,
for $1 \le i \le N$,
then
$$
    \P\{ |X| > 2t \} \approx \frac{1}{N} \sum_{i : |e_i| > 2t} p_i .
$$
So, to obtain a per-trait estimate of $\alpha$,
we plot $\log \frac{1}{N} \sum_{i : |e_i| > t} p_i$ against $\log t$
for a sequence of values of $t$,
and fit a linear model to the tail
(which we take to be all values of $t$ for which the number of SNPs with absolute effect above $t$
is between 20 and 10\% of all SNPs).
Examples are shown in figure~\ref{fig:example_snps}.

\begin{figure}
    \begin{center}
    \includegraphics{snp_effects/examples}
    \end{center}
    \caption{
        TODO: cut this down to just 2 SNPs and move this figure to the supplement; change 'x' to 't' in axis labels.
        Tail distributions of frequency-weighted effects
        for twelve randomly chosen phenotypes.
        In each, for a phenotype with $N$ SNPs,
        effect sizes $e_i$ and frequencies $p_i$,
        the black line shows
        $\log \frac{1}{N} \sum_{i : |e_i| > t} p_i$ (vertical axis)
        against $\log t$ (horizontal axis);
        the red line shows the least-squares linear fit;
        each plot only covers the 10\% largest-effect SNPs.
        Legends show the coefficients of the red line; the slope on `x' is $\alpha$,
        the estimated exponent;
        ``num snps'' is the number of SNPs for the phenotype with $p$-value less than $10^{-8}$,
        and ``num cases'' is the reported number of the rougly 360,000 individuals
        recorded as having the listed phenotype.
        \label{fig:example_snps}
    }
\end{figure}

We applied this procedure to results from models fit on both sexes,
after removed phenotypes related to employment, drug treatments, and diet.
This left us with 1,108 phenotypes for which which could estimate the tail exponent.
For the main results,
we then restricted to phenotypes having non-missing numbers of ``cases'' and ``controls''
and a sample size of at least 300,000 (i.e., not more than 61,194 of 361,194 missing phenotypes).
We also removed
the 65 remaining phenotypes that are common (more than 10,000 cases),
as these often showed very different patterns
(e.g., effect size distributions).
This left us with 695 phenotypes;
plots including the 413 ``filtered'' phenotypes are provided in
Supplementary Figure~\ref{fig:unfiltered_hist}.

Resulting values of the tail exponent $\alpha$ are shown in Figure~\ref{fig:exponent_hist}.
Estimated tail exponents range from about 1 to 2.5,
and 28.9\% had estimated values less than 1.5.
A concern could be that ``noisier'' phenotypes,
for which estimated effect sizes are less reliable,
might lead to larger estimated tail exponents;
for this reasons, we looked for an association between
tail exponent and two proxies for power,
number of SNPs (i.e., number of SNPs with $p$-value less than $10^{-8}$)
and number of cases.
Interestingly, the result in nonmonotonic,
with exponents closer to $\alpha=2$ for phenotypes with around 1000 cases,
and lower exponents for rarer \emph{and} more common phenotypes.
(A similar pattern is seen for ``number of SNPs'',
but this may be due to association with number of cases.)
This may indicate a statistical artifact,
or it may be a result of biological differences in genetic architecture
between more and less common phenotypes.

@@

We generally follow the notation of \citet{barton2017infinitesimal},
although in a simpler situation,
as we consider only the haploid offspring of a single set of haploid parents
rather than full pedigrees,
and ignore non-genetic contributions to the trait;
so, below we say ``trait'' to mean ``genetic contribution to the trait''.

Suppose that the focal individual's trait is
$$
    Z_M = \sum_{\ell=1}^M \eta_\ell / M^{1/\alpha},
$$
where $\eta_\ell$ is the contribution of the allele carried at the $\ell^\text{th}$ locus.
The effects of the alleles carried by the two parents
are $\eta_\ell^{[1]}$ and $\eta_\ell^{[2]}$, respectively,
and their traits are
$Z_M^{[1]} = \sum_\ell \eta_\ell^{[1]} / M^{1/\alpha}$ and
$Z_M^{[2]} = \sum_\ell \eta_\ell^{[2]} / M^{1/\alpha}$.
We assume the loci are unlinked and inheritance is fair, so
letting $X_\ell$ be a family of Bernoulli(1/2) random variables that are independent of everything else,
$$
    \eta_\ell = X_\ell \eta_\ell^{[1]} + (1 - X_\ell) \eta_\ell^{[1]} .
$$

@@

Is there a converse? In other words,
does independence of segregation noise imply a Gaussian distribution of trait values?
It is not clear to us in how much generality such a statement might be made,
but we can at least prove the following.
Divide the parental loci into groups according to the results of inheritance,
defining
\begin{align*}
    U_M^{[i]} = \sum_{\ell=1}^M X_\ell \eta_\ell^{[i]} 
    \qquad \text{ and } \qquad
    V_M^{[i]} = \sum_{\ell=1}^M (1-X_\ell) \eta_\ell^{[i]} ,
\end{align*}
so that 
$Z_M^{[1]} = U_M^{[1]} + V_M^{[1]}$ and
$Z_M^{[2]} = U_M^{[2]} + V_M^{[2]}$, so
\begin{align*}
    Z_M = U_M^{[1]} + V_M^{[2]} \qquad \text{ and } \qquad
    R_M = \frac{U_M^{[1]} - V_M^{[1]} - U_M^{[2]} + V_M^{[2]}}{ 2 } .
\end{align*}
It turns out that independence of the portions inherited and not inherited by the offspring
combined with independence of segregation noise and midparent value
implies a Gaussian distribution:

\begin{lem}\label{lem:parental_contribs}
    Suppose that $\{\eta_\ell^{[i]}\}$ are independent for all $\ell$ and $i$,
    that the distribution of $\eta_\ell^{[i]}$ only depends on $\ell$,
    and that $\{X_\ell\}$ are independent of $\{\eta_\ell^{[i]}\}$.
    If the segregation noise is independent of the midparent value
    (i.e., $R_M$ is independent of $\bar Z_M$)
    then $R_M$ is Gaussian.
\end{lem}

\begin{proof}[Proof of Lemma~\ref{lem:parental_contribs}]
    Note that if inheritance (the $X_\ell$) is independent of the per-locus effects ($\eta_\ell^{[i]}$),
    the latter are all independent of each other,
    and all effects at a given locus are drawn from the same distribution,
    then the inherited allele, 
    $\eta_\ell^{[1]} X_\ell + \eta_\ell^{[2]} (1-X_\ell)$
    is independent of (and has the same distribution as)
    the non-inherited allele,
    $\eta_\ell^{[1]} (1-X_\ell) + \eta_\ell^{[2]} X_\ell$
    -- they are simply two independent draws from the distribution of effects at locus $\ell$.
    Therefore, the sum of inherited alleles, $U_M^{[1]} + V_M^{[2]}$
    are independent of the non-inherited alleles, $U_M^{[2]} + V_M^{[1]}$.

    Now, the Kac-Bernstein theorem states that if two independent random variables
    have the property that if their sum and difference are independent of each other,
    then the random variables have a joint Gaussian distribution
    (\citet{kac1939characterization,bernstein1941property},
    but see \citet{kagan1973characterization} for many much more general results).
    Since $R_M$ is (half) the difference and $(Z^{[1]} + Z^{[2]})/2$ is (half) the sum
    of $U_M^{[1]} + V_M^{[2]}$ and $U_M^{[2]} + V_M^{[1]}$,
    both are therefore jointly Gaussian.
\end{proof}

Since models can be easily set up for which $Z$ is not Gaussian,
the content of this is that for such models, independence of segregation noise
is not likely a good assumption.
Note that the proof does not depend on independence of the $X_\ell$:
indeed, alleles may be inherited together,
although in practice this would quickly create dependencies between nearby allelic effects.
Also note that it is \emph{not} usually true that the components --
for instance, $U_M^{[1]}$ and $U_M^{[2]}$ --
are independent of each other.

\comment{remove or shorten the following?}
It is clear in the Gaussian case
that $U^{[1]}$ and $V^{[1]}$ are asymptotically independent.
More generally, independence depends on how the model is set up.
If the allelic effects $\eta_\ell^{[i]}$ are i.i.d.,
then even for finite $M$, the parental contributions
$(U_M^{[1]}, V_M^{[1]}, U_M^{[2]}, V_M^{[2]})$ are independent.
If the distribution of $\eta_\ell^{[i]}$ is chosen to obtain a non-Gaussian limit,
then Lemma~\ref{lem:parental_contribs} implies that $R$ cannot be independent of the parental traits.

However, the presence of a large effect polymorphic allele
makes $(U^{[1]}, V^{[1]}, U^{[2]}, V^{[2]})$ dependent:
consider the extreme case where $\eta^{[i]}_\ell$ are i.i.d.{} with variance 1 for $\ell \ge 2$,
while $\eta^{[i]}_1 = \pm 10$ with probability 1/2 each,
and $Z^{[1]} = \eta_1^{[1]} + \sum_{\ell=2}^M \eta_\ell^{[1]} / \sqrt{M}$.
Then in the limit $U^{[i]}$ and $V^{[i]}$ are Gaussian with mean either 0 or $\pm 10$ and variance 1/2,
and are highly dependent: $|U^{[i]}| \approx 10$ or $|V^{[i]}| \approx 10$ but not both.
(Segregation noise, meanwhile, depends strongly on parental trait values:
if $Z^{[1]} \approx Z^{[2]}$ then most likely
$\eta_1^{[1]} = \eta_1^{[2]}$ and so $R$ is Normal with variance 1/2;
but in the other case then $R$ has equal probabilities of being close to $+5$ or $-5$.)

@@

Let $X$ be a population of size $N$, recorded as the empirical distribution on phenotypes
\begin{align*}
    X = \frac{1}{N} \sum_{i=1}^N \delta_{x_i} .
\end{align*}
So, $X$ is a point measure of total mass 1 on $\R$.
This evolves as a Moran process under the infinitesimal model
with death rates depending on state $x$.
Concretely, an individual whose phenotype is $x$
dies at rate $1 + \mu(x)$
and is replaced by a new individual of phenotype
\begin{align*}
  \frac{y + z}{2} + \xi ,
\end{align*}
where $y$ and $z$ are independent draws from $X$
and $\xi$ is an independent draw from the segregation noise distribution.
Let $p_\xi$ denote the distribution of $\xi$,
and $\nu(X)$ denote the distribution of $\frac{y + z}{2} + \xi$.
This has generator
\begin{align*}
    \mL \Phi(X)
    &=
    \int \int \int \int
    \left( \Phi\left( X + \frac{1}{N} \delta_{(y+z)/2 + \xi} - \frac{1}{N} \delta_x \right) - \Phi(X) \right)
    (1 + \mu(x)) X(dx) X(dy) X(dz) p_\xi(d\xi) \\
    &=
    \int \int
    \left( \Phi\left(X + \frac{1}{N} \delta_u - \frac{1}{N} \delta_x\right) - \Phi(X) \right)
    (1 + \mu(x)) X(dx) \nu(X)(du) .
\end{align*}

@@

More generally, if $\Phi(X) = \prod_i \ip{\phi_i}{X}$ then
\begin{align*}
    \mL \Phi(X)
    &=
    \frac{1}{N} \sum_i \left( 
        \ip{1+\mu}{X} \ip{\phi_i}{\nu(X)}  - \ip{\phi_i}{(1+\mu)X}
    \right)
    \prod_{j \neq i} \ip{\phi_j}{X} 
    + O(1/N^2) .
\end{align*}
This implies that after rescaling time by $N$
and suitible other caveats,
the limiting process is a deterministic flow in the direction of 
$\nu(X) \ip{1+\mu}{X} - (1+\mu)X$.

>>>>

% Intro:
%  - the infinitesimal model is useful because it is a "trait-only" model: it allows for predictions and modeling without knowing the underlying genetics.
%  - Its key properties are: (a) independence of seg noise, and (b) Gaussian-ness.
%  - It is basically a CLT, and in particular, (b) derives from a the CLT deriving from similar effect sizes.
%  - Motivating Q: there are other CLTs; can we generalize (b) to non-Gaussian effects, while keeping independence of seg variation?
% 
% Data motivation:
%  First, to see if this is worth asking we look at some data
%  ... and find that yeah, maybe effect sizes don't fall into the Gaussian domain of attraction.
% 
% First answer to the motivating question:  "No"
% Non-Gaussian effects (sometimes) imply nonindependence of transmission noise:
%   In at least some cases, natural non-gaussian models produce nonindependence:
%   1. all loci are uniformly small except there's one big one
%   2. loci are Cauchy
%   3. (is there a simple statement we can make more generally?)
% 
% Second answer, weaker but also "no":
%  - Observation: Let A and B be the contributions to parent phenotype, with A inherited by the offspring and B not. If A and B are independent, and seg noise is independent of parent phenotype, then A and B are Gaussian.
%  - When would A and B be asymptotically independent? One case is the rare-mutation limit that obtains a "thinned Poisson process" limit for A and B (I think: if all allele frequencies are at frequency O(1/M)?)
%  - (I have other thoughts to be sorted out here)
%  - Open Q (maybe?): is there any non-Gaussian model where transmission noise is independent of parental traits?
% 
% Simulations: let's see how this stuff works in practice
%  - neutral simulations
%  - also, stabilizing selection with theoretical motivation for form of stabilizing selection
%  - plot: (A) trait mean over time and trait distribution at the end
%  - plot: (B) (distrn of seg noise), (seg noise vs midparent), and (Radon-Nicodym deriv of conditional seg noise for a few values of midparent)
% 
% If we remove 'big loci' is what's left the infinitesimal model?
%  - people do this in practice (animal breeding papers)
%  - theory:  form of limiting conditional distribution
%  - theory: something about independence  (?? if we get there ??)
%  - plots (A) & (B) above from simulation, but for only small effects

@@

\begin{abstract}
    The infinitesimal model of quantitative genetics
    relies on the Central Limit Theorem to stipulate that
    under additive models of quantitative traits
    determined by many loci having similar effect size,
    the difference between an offspring's genetic trait component
    and the average of their two parents' genetic trait components
    is Normally distributed and independent of the parents' values.
    Here, we investigate how the assumption of similar effect sizes
    affects the model:
    in particular, if the tail of the effect size distribution is polynomial
    with exponent $\alpha < 2$, then sums of effects
    should be well-approximated by a ``stable distribution'',
    and find tail exponents between 1 and 2 in 
    effect sizes estimated by genome-wide association studies
    of many human disease-related traits.
    We show that the independence of offspring trait deviations from parental averages
    in many cases implies a Gaussian distribution,
    suggesting that non-Gaussian models of trait evolution must explicitly
    track the underlying genetics, at least for loci of large effect.
    We also characterize possible limiting trait distributions
    of the infinitesimal model with infinitely divisible noise distributions,
    and compare our results to simulations.
\end{abstract}

@@

Suppose that the focal individual's trait is
$$
    Z_M = \sum_{\ell=1}^M \eta_\ell / M^{1/\alpha},
$$
where $\eta_\ell$ is the contribution of the allele carried at the $\ell^\text{th}$ locus.
The effects of the alleles carried by the two parents
are $\eta_\ell^{[1]}$ and $\eta_\ell^{[2]}$, respectively,
and their traits are
$Z_M^{[1]} = \sum_\ell \eta_\ell^{[1]} / M^{1/\alpha}$ and
$Z_M^{[2]} = \sum_\ell \eta_\ell^{[2]} / M^{1/\alpha}$.
We assume the loci are unlinked and inheritance is fair, so
letting $X_\ell$ be a family of Bernoulli(1/2) random variables that are independent of everything else,
$$
    \eta_\ell = X_\ell \eta_\ell^{[1]} + (1 - X_\ell) \eta_\ell^{[2]} .
$$
What we will call the \emph{transmission noise}, $R_M$,
is the offspring's trait minus the midparent value, $\bar Z_M$:
\begin{align*}
    \bar Z_M &= \frac{Z_M^{[1]} + Z_M^{[2]}}{2} \\
    R_M &= Z_M - \bar Z_M \\
        &= \sum_{\ell=1}^M
            \frac{
                X_\ell (\eta_\ell^{[1]} - \eta_\ell^{[2]})
                + (1 - X_\ell) (\eta_\ell^{[2]} - \eta_\ell^{[1]})
            }{ 2 } .
\end{align*}

@@

It is mathematically intriguing to speculate what effect a power law effect size distribution
for would have on the induced evolutionary dynamics.
But, is there any evidence that non-Gaussian transmission noise is important in practice?
In this paper we focus primarily on the theory,
but first take a brief and imperfect look at what some data have to say on the topic.
To do this, we downloaded the GWAS results provided by~\citet{biobankSNPs}
for 7,221 phenotypes from 361,194 human individuals from the March 2018 release of the UK Biobank.
After filtering (see details below) we were left with 695 binary illness-related phenotypes.
Taken at face value, these provide for each phenotype a set of estimates of the additive effect
of the alternative allele at a set of single nucleotide polymorphisms (SNPs)
on the phenotype relative to the reference allele.
The phenotypes we consider were all binary and coded as 0/1,
so estimates could be thought of as additive adjustments to the risk
(so, effects are \emph{not} on a logit scale).
These estimates were obtained as the coefficient for the SNP in a simple linear model
fit to each phenotype for each SNP,
with the first 20 principal components and all combinations of age, age squared, and inferred sex as covariates;
we considered effects of those SNPs with a reported $p$-value less than $10^{-8}$.
There are a great many possible issues with these data;
however, the results seem consistent across many SNPs and phenotypes,
and we know of no reason that possible methodological artifacts
(e.g., confounding by uncorrected variables)
would specifically affect the decay of the tail of effect sizes.

Since the transmission noise is a sum of terms of the form
$$
    (X_\ell - 1/2) \eta_\ell^{[1]} + (1/2 - X_\ell) \eta_\ell^{[2]},
$$
where $\eta_\ell^{[i]}$ is the effect of the allele in parent $i$ at locus $\ell$,
the appropriate central limit theorem for the sum is determined by
the tail behavior of $|(X - 1/2) \eta| = |\eta|/2$,
where $X$ is Bernoulli(1/2) and $\eta$ is the effect
of a uniformly chosen allele at a uniformly chosen locus.
So, we'd like to know
if $\P\{|X|>2t\}$ is proportional to $t^{-\alpha}$,
i.e., the value of
$$
    \lim_{t \to \infty} \frac{ \log \P\{ |X| > 2 t \} }{ \log t } = \alpha ,
$$
if the limit exists.
Let $p_\ell$ be the minor allele frequency at the $\ell^\text{th}$ locus;
since the data we have assigns effect 0 to the major allele,
$\eta_\ell = 0$ with probability $1-p_\ell$.
Therefore, if we have estimated effects for $M$ SNPs
with minor allele frequency $p_\ell$ and effect size $e_\ell$ for the minor allele,
for $1 \le \ell \le M$,
then
$$
    \P\{ |\eta| > 2t \} \approx \frac{1}{M} \sum_{\ell : |e_\ell| > 2t} p_\ell .
$$
So, to obtain a per-trait estimate of $\alpha$,
we plot $\log \frac{1}{M} \sum_{\ell : |e_\ell| > t} p_\ell$ against $\log t$
for a sequence of values of $t$,
and fit a linear model to the tail
(which we take to be all values of $t$ for which the number of SNPs with absolute effect above $t$
is between 20 and 10\% of all SNPs).
Examples are shown in Figure~\ref{fig:example_snps}.

\begin{figure}
    \begin{center}
    \includegraphics{snp_effects/examples}
    \end{center}
    \caption{
        % TODO: cut this down to just 2 SNPs and move this figure to the supplement; change 'x' to 't' in axis labels.
        Tail distributions of frequency-weighted effects
        for twelve randomly chosen phenotypes.
        In each, for a phenotype with $N$ SNPs,
        effect sizes $e_i$ and frequencies $p_i$,
        the black line shows
        $\log \frac{1}{N} \sum_{i : |e_i| > t} p_i$ (vertical axis)
        against $\log t$ (horizontal axis);
        the red line shows the least-squares linear fit;
        each plot only covers the 10\% largest-effect SNPs.
        Legends show the coefficients of the red line; the slope on `x' is $\alpha$,
        the estimated exponent;
        ``num snps'' is the number of SNPs for the phenotype with $p$-value less than $10^{-8}$,
        and ``num cases'' is the reported number of the roughly 360,000 individuals
        recorded as having the listed phenotype.
        \label{fig:example_snps}
    }
\end{figure}

We applied this procedure to results from models fit on both sexes,
after removing phenotypes related to non-medical traits
(e.g., employment, drug treatments, and dietary choices).
This left us with 1,108 phenotypes for which we had enough data (e.g., enough significant SNPs)
to estimate the tail exponent.
We then restricted to phenotypes having non-missing numbers of ``cases'' and ``controls''
and a sample size of at least 300,000 (i.e., not more than 61,194 of 361,194 missing phenotypes),
and removed the 65 remaining phenotypes that are common (more than 10,000 cases),
as these often showed very different patterns
(e.g., effect size distributions).
This left us with 695 phenotypes;
plots including the 413 phenotypes removed by these last steps are provided in
Supplementary Figure~\ref{fig:unfiltered_hist}.

Resulting values of the tail exponent $\alpha$ are shown in Figure~\ref{fig:exponent_hist}.
Estimated tail exponents range from about 1 to 2.5,
and 28.9\% had estimated values less than 1.5.
A concern is that ``noisier'' phenotypes,
for which estimated effect sizes are less reliable,
might lead to larger estimated tail exponents;
for this reason, we looked for an association between
tail exponent and two proxies for power,
number of SNPs (i.e., number of SNPs with $p$-value less than $10^{-8}$)
and number of cases.
Interestingly, the relationship is nonmonotonic,
with exponents closer to $\alpha=2$ for phenotypes with around 1000 cases,
and lower exponents for both rarer and more common phenotypes.
(A similar pattern is seen for ``number of SNPs'',
but this may be due to association with power and hence number of cases.)
This may indicate a statistical artifact,
or it may be a result of biological differences in genetic architecture
between more and less common phenotypes.

@@

@@

Perhaps surprisingly, there is a converse --
in at least some situations,
independence of transmission noise implies a Gaussian distribution.
To see this,
divide the parental loci into groups according to the results of inheritance,
defining
\begin{align*}
    U_M^{[1]} &= \sum_{\ell=1}^M X_\ell \eta_\ell^{[1]}  ,
    \qquad \qquad
    V_M^{[1]} = \sum_{\ell=1}^M (1-X_\ell) \eta_\ell^{[1]} ,  \\
    U_M^{[2]} &= \sum_{\ell=1}^M (1-X_\ell) \eta_\ell^{[2]} ,
    \qquad \qquad
    V_M^{[2]} = \sum_{\ell=1}^M X_\ell \eta_\ell^{[2]}  ,
\end{align*}
so that 
$Z_M^{[1]} = U_M^{[1]} + V_M^{[1]}$ and
$Z_M^{[2]} = U_M^{[2]} + V_M^{[2]}$, while
$Z_M = U_M^{[1]} + U_M^{[2]}$ is the transmitted portion of the parental genomes,
and we can define $V_M = V_M^{[1]} + V_M^{[2]}$ to be the untransmitted portion.
Furthermore,
\begin{align*}
    \bar Z_M = \frac{Z_M + V_M}{2}
    \qquad \text{ and } \qquad
    R_M = \frac{Z_M - V_M}{ 2 } .
\end{align*}

Now, note that if inheritance (the $X_\ell$) is independent of the per-locus effects ($\eta_\ell^{[i]}$),
the latter are all independent of each other,
and all effects at a given locus are drawn from the same distribution,
then the inherited allele, 
$\eta_\ell^{[1]} X_\ell + \eta_\ell^{[2]} (1-X_\ell)$
is independent of (and has the same distribution as)
the non-inherited allele,
$\eta_\ell^{[1]} (1-X_\ell) + \eta_\ell^{[2]} X_\ell$
-- they are simply two independent draws from the distribution of effects at locus $\ell$.
Therefore, the inherited alleles that sum to $Z_M$
are independent of the non-inherited alleles that sum to $V_M$.

The Kac-Bernstein theorem states that if two independent random variables
have the property that if their sum and difference are independent of each other,
then the random variables have a joint Gaussian distribution
(\citet{kac1939characterization,bernstein1941property},
and see \citet{kagan1973characterization} for many much more general results).
Since $R_M$ is (half) the difference and $\bar Z_M$ is (half) the sum
of $Z_M$ and $V_M$, and $Z_M$ and $V_M$ are independent,
if we assume that also $R_M$ and $\bar Z_M$ are independent,
then both are therefore jointly Gaussian.

We have shown the following:

\begin{prop}\label{prop:parental_contribs}
    Suppose that $\{\eta_\ell^{[i]}\}$ are independent for all $\ell$ and $i$,
    that the distribution of $\eta_\ell^{[i]}$ only depends on $\ell$,
    and that $\{X_\ell\}$ are independent of $\{\eta_\ell^{[i]}\}$.
    If the transmission noise $R_M$ is independent of the midparent value $\bar Z_M$,
    then $R_M$ and $\bar Z_M$ are jointly Gaussian.
\end{prop}

% A similar statement holds for the offspring of two parents in a pedigree,
% using the Skitovich-Darmois theorem \citet{kagan1973characterization},
% except that the portion of the genome shared in common by the parents is unconstrained.

Since models can be easily set up for which trait distributions are not Gaussian,
the implication of this is that for such models, independence of transmission noise
is not likely a good assumption.
Note that the proof does not depend on independence of the $X_\ell$:
indeed, alleles may be inherited together,
although in practice this would quickly create dependencies between nearby allelic effects.
Also note that it is \emph{not} usually true that the components --
for instance, $U_M^{[1]}$ and $U_M^{[2]}$ --
are independent of each other.

The presence of a large effect polymorphic allele
makes $R_M$ and $\bar Z_M$ dependent:
consider the extreme case where $\eta^{[i]}_\ell$ are i.i.d.{} with variance 1 for $\ell \ge 2$,
while $\eta^{[i]}_1 = \pm 10$ with probability 1/2 each,
and $Z^{[i]} = \eta_1^{[i]} + \sum_{\ell=2}^M \eta_\ell^{[i]} / \sqrt{M}$.
Then in the limit $R_M$ depends strongly on parental trait values:
if $Z^{[1]} \approx Z^{[2]}$ and so $\bar Z \approx \pm 10$ then most likely
$\eta_1^{[1]} = \eta_1^{[2]}$ and so $R$ is Normal with variance 1/2;
but if $\eta_1^{[1]} \neq \eta_1^{[2]}$ (and so $\bar Z \approx 0$)
then $R$ is close to $+5$ or $-5$ with equal probability.

@@

Let $X$ be a population of size $N$, recorded as the empirical distribution on phenotypes
\begin{align*}
    X = \frac{1}{N} \sum_{i=1}^N \delta_{x_i} .
\end{align*}
So, $X$ is a point measure of total mass 1 on $\R$.
This evolves as a Moran process under the infinitesimal model
with death rates depending on state $x$.
Concretely, an individual whose phenotype is $x$
dies at rate $1 + \mu(x)$
and is replaced by a new individual of phenotype
\begin{align*}
  \frac{y + z}{2} + \xi ,
\end{align*}
where $y$ and $z$ are independent draws from $X$
and $\xi$ is an independent draw from the transmission noise distribution.
Let $p_\xi$ denote the distribution of $\xi$,
and $\nu(X)$ denote the distribution of $\frac{y + z}{2} + \xi$.
This has generator
\begin{align*}
    \mL \Phi(X)
    &=
    \int \int \int \int
    \left( \Phi\left( X + \frac{1}{N} \delta_{(y+z)/2 + \xi} - \frac{1}{N} \delta_x \right) - \Phi(X) \right)
    (1 + \mu(x)) X(dx) X(dy) X(dz) p_\xi(d\xi) \\
    &=
    \int \int
    \left( \Phi\left(X + \frac{1}{N} \delta_u - \frac{1}{N} \delta_x\right) - \Phi(X) \right)
    (1 + \mu(x)) X(dx) \nu(X)(du) .
\end{align*}

@@

More generally, if $\Phi(X) = \prod_i \ip{\phi_i}{X}$ then
\begin{align*}
    \mL \Phi(X)
    &=
    \frac{1}{N} \sum_i \left( 
        \ip{1+\mu}{X} \ip{\phi_i}{\nu(X)}  - \ip{\phi_i}{(1+\mu)X}
    \right)
    \prod_{j \neq i} \ip{\phi_j}{X} 
    + O(1/N^2) .
\end{align*}
This implies that after rescaling time by $N$
and suitable other caveats,
the limiting process is a deterministic flow in the direction of 
$\nu(X) \ip{1+\mu}{X} - (1+\mu)X$.
